---
title: "Seismic Enhanced Features Guide"
author: "seismicGWAS Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Seismic Enhanced Features Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette demonstrates the enhanced features added to the seismicGWAS package, including:

1. Non-linear modeling
2. Multi-omics integration
3. Hierarchical modeling
4. Causal inference (MR)
5. Enhanced influence analysis with stability
6. Trajectory/dynamic analysis
7. Batch effect handling
8. Uncertainty quantification
9. Simulation and benchmarking

# Installation

```{r, eval=FALSE}
# Install from GitHub
devtools::install_github("Charlesyip2016/seismic")
library(seismicGWAS)
```

# 1. Non-linear Modeling

Test for non-linear relationships using GAM:

```{r, eval=FALSE}
# Standard linear model
results_linear <- get_ct_trait_associations(sscore, magma_data)

# GAM model to detect non-linearity
results_gam <- get_ct_trait_associations_nonlinear(
  sscore, magma_data,
  model_type = "gam",
  gam_k = 10
)

# Compare AIC - if delta > 2, non-linearity is present
results_gam[nonlinear == TRUE]
```

Test for interaction effects:

```{r, eval=FALSE}
# Test p × r interactions
interaction_results <- test_specificity_interactions(
  sce, magma_data,
  ct_label_col = "cell_type"
)

# Significant interactions
interaction_results[FDR_interaction < 0.05]
```

# 2. Multi-omics Integration

Combine expression, ATAC, eQTL, and colocalization data:

```{r, eval=FALSE}
# Calculate gene activity from ATAC-seq
atac_scores <- calc_gene_activity_from_atac(
  atac_sce, 
  peak_to_gene_links,
  ct_label_col = "cell_type"
)

# Prepare eQTL scores
eqtl_scores <- prepare_eqtl_scores(
  eqtl_results,
  score_type = "pvalue"
)

# Prepare colocalization scores
coloc_scores <- prepare_coloc_scores(
  coloc_results,
  pp_threshold = 0.5
)

# Create composite scores
composite_scores <- calc_composite_specificity(
  expr_specificity = sscore,
  atac_specificity = atac_scores,
  eqtl_scores = eqtl_scores,
  coloc_scores = coloc_scores,
  weights = c(expr = 1.0, atac = 0.5, eqtl = 0.5, coloc = 0.8)
)

# Test associations with composite scores
results_composite <- get_ct_trait_associations(composite_scores, magma_data)
```

# 3. Hierarchical Modeling

Use cell type hierarchy to improve rare cell type estimates:

```{r, eval=FALSE}
# Define hierarchy
cell_hierarchy <- data.frame(
  cell_type = c("B_cell", "T_CD4", "T_CD8", "NK", "Monocyte"),
  parent_type = c("Lymphoid", "T_cell", "T_cell", "Lymphoid", "Myeloid")
)

# Hierarchical association analysis
hier_results <- get_hierarchical_associations(
  sscore, magma_data, cell_hierarchy,
  tau_prior = 1.0
)

# Check shrinkage factors
hier_results[order(shrinkage_factor)]

# Calculate specificity with empirical Bayes shrinkage
sscore_shrunk <- calc_specificity_with_shrinkage(
  sce,
  ct_label_col = "cell_type",
  cell_hierarchy = cell_hierarchy,
  min_cells_shrinkage = 50
)
```

# 4. Causal Inference

Perform Mendelian Randomization:

```{r, eval=FALSE}
# MR analysis for a cell type
mr_results <- perform_mendelian_randomization(
  eqtl_data, gwas_data,
  cell_type = "B_cell",
  f_threshold = 10,
  pval_threshold = 5e-8
)

# Test for pleiotropy
pleiotropy <- test_pleiotropy_egger(
  eqtl_data, gwas_data,
  cell_type = "B_cell"
)

# Flag genes with pleiotropy
mr_results[pleiotropy$pleiotropy_detected == TRUE]

# Mediation analysis
mediation_result <- perform_mediation_analysis(
  snp_dosage, gene_expression, trait_values,
  bootstrap = TRUE,
  n_bootstrap = 1000
)

# Combine MR with specificity
causal_scores <- calc_causal_specificity_scores(
  mr_results, sscore,
  mr_pval_threshold = 0.05
)
```

# 5. Enhanced Influence Analysis

Identify influential genes with stability assessment:

```{r, eval=FALSE}
# Standard DFBETAS
inf_genes_basic <- find_inf_genes("B_cell", sscore, magma_data)

# With stability selection
inf_genes_stable <- find_inf_genes_stable(
  "B_cell", sscore, magma_data,
  n_bootstrap = 100,
  sampling_fraction = 0.8
)

# Core drivers (high stability)
inf_genes_stable[stability_category == "high"]

# Leave-k-out analysis
lko_results <- leave_k_out_analysis(
  "B_cell", sscore, magma_data,
  k_percent = 5,
  n_iterations = 50
)

# Check stability
lko_results$stability_ratio  # Higher = more stable

# Pathway-level influence
gene_groups <- list(
  immune_response = c("CD19", "CD79A", "MS4A1"),
  apoptosis = c("BCL2", "BAX", "CASP3")
)

pathway_influence <- group_influence_analysis(
  "B_cell", sscore, magma_data, gene_groups
)
```

# 6. Trajectory and Dynamic Analysis

Analyze stage-specific effects:

```{r, eval=FALSE}
# Calculate trajectory-specific scores
stage_specificity <- calc_trajectory_specificity(
  sce,
  pseudotime_col = "pseudotime",
  ct_label_col = "cell_type",
  n_windows = 5
)

# Test stage-specific associations
stage_results <- test_trajectory_associations(
  stage_specificity, magma_data
)

# Find stage-specific driver genes
stage_drivers <- find_stage_specific_drivers(
  cell_type = "B_cell",
  stage = "stage2",
  stage_specificity, magma_data
)

# Compare across stages
stage_comparison <- compare_stage_associations(
  "B_cell", stage_specificity, magma_data
)

# Test for heterogeneity
stage_comparison[heterogeneity_pval < 0.05]

# Dynamic expression modeling
dynamic_results <- test_dynamic_associations(
  sce, 
  pseudotime_col = "pseudotime",
  ct_label_col = "cell_type",
  magma_data
)
```

# 7. Batch Effect Handling

Correct for batch effects:

```{r, eval=FALSE}
# Define batch information
batch_info <- data.frame(
  cell_type = colnames(sscore),
  batch = c(rep("Batch1", 5), rep("Batch2", 5))
)

# Correct batch effects
sscore_corrected <- correct_batch_effects(
  sscore, batch_info,
  method = "mean_center"
)

# Stratified analysis
sscore_list <- list(
  Batch1 = sscore_batch1,
  Batch2 = sscore_batch2
)

batch_meta <- stratified_batch_analysis(
  sscore_list, magma_data,
  meta_method = "fixed"
)

# Check heterogeneity
batch_meta$meta_results[heterogeneity_pval < 0.05]

# Add covariates
covariates <- data.frame(
  gene = genes,
  gc_content = gc_values,
  gene_length = lengths
)

results_adj <- get_associations_with_covariates(
  sscore, magma_data, covariates,
  covariate_cols = c("gc_content", "gene_length")
)

# Cross-dataset replication
replication <- assess_cross_dataset_replicability(
  sscore_dataset1, sscore_dataset2, magma_data,
  fdr_threshold = 0.05
)

print(replication$replication_rate)
```

# 8. Uncertainty Quantification

Get confidence intervals and assess calibration:

```{r, eval=FALSE}
# Bootstrap confidence intervals
results_ci <- get_associations_with_uncertainty(
  sscore, magma_data,
  n_bootstrap = 1000,
  ci_level = 0.95
)

# Check which CIs exclude 0
results_ci[ci_lower > 0]

# Assess p-value calibration
calibration <- assess_pvalue_calibration(results)

# Check genomic inflation
print(paste("Lambda:", round(calibration$lambda, 3)))

# QQ plot
plot(calibration$expected, calibration$observed,
     xlab = "Expected -log10(p)",
     ylab = "Observed -log10(p)",
     main = "QQ Plot")
abline(0, 1, col = "red")
```

# 9. Simulation and Benchmarking

Test methods on simulated data:

```{r, eval=FALSE}
# Generate simulated data
sim_sc <- simulate_sc_data(
  n_genes = 5000,
  n_cells_per_type = c(100, 200, 150, 80),
  cell_type_names = c("TypeA", "TypeB", "TypeC", "TypeD"),
  n_driver_genes = 50,
  driver_effect_size = 2.0,
  seed = 42
)

# Calculate specificity
sscore_sim <- calc_specificity(sim_sc$sce, ct_label_col = "cell_type")

# Generate GWAS
gwas_sim <- simulate_gwas_data(
  rownames(sscore_sim),
  sim_sc$driver_genes,
  sscore_sim,
  effect_size = 0.5
)

# Benchmark methods
benchmark_results <- benchmark_seismic(
  sim_sc, gwas_sim,
  methods = c("linear", "gam", "hierarchical"),
  fdr_threshold = 0.05
)

# View performance
print(benchmark_results$performance)

# Generate full simulation suite
scenarios <- generate_simulation_suite(
  n_scenarios = 10,
  vary_params = c("sample_size", "effect_size", "noise"),
  seed = 42
)

# Run benchmarks on all scenarios
all_results <- lapply(scenarios, function(scenario) {
  benchmark_seismic(
    scenario$sc_data,
    scenario$gwas_data,
    methods = c("linear", "hierarchical")
  )
})
```

# Complete Workflow Example

Here's an end-to-end analysis:

```{r, eval=FALSE}
# 1. Load data
library(seismicGWAS)
data("tmfacs_sce_small")
data("t2d_magma")

# 2. Calculate enhanced specificity (with shrinkage for rare cells)
sscore <- calc_specificity_with_shrinkage(
  tmfacs_sce_small,
  ct_label_col = "cluster_name",
  min_cells_shrinkage = 30
)

# 3. Test associations with uncertainty
results <- get_associations_with_uncertainty(
  sscore, t2d_magma,
  n_bootstrap = 500
)

# 4. Test non-linearity for top cell types
top_cts <- results[FDR < 0.05, cell_type]
nonlinear_test <- get_ct_trait_associations_nonlinear(
  sscore[, top_cts], t2d_magma,
  model_type = "gam"
)

# 5. Find influential genes with stability
for (ct in top_cts) {
  inf_genes <- find_inf_genes_stable(
    ct, sscore, t2d_magma,
    n_bootstrap = 100
  )
  
  # Save high-stability drivers
  core_drivers <- inf_genes[stability_category == "high"]
  write.csv(core_drivers, 
           paste0("drivers_", gsub(" ", "_", ct), ".csv"),
           row.names = FALSE)
}

# 6. Assess calibration
calibration <- assess_pvalue_calibration(results)
message("Genomic inflation factor: ", round(calibration$lambda, 3))

# 7. Summary report
summary_report <- data.frame(
  n_cell_types = length(unique(results$cell_type)),
  n_significant = sum(results$FDR < 0.05),
  n_nonlinear = sum(nonlinear_test$nonlinear, na.rm = TRUE),
  lambda = calibration$lambda
)

print(summary_report)
```

# Session Info

```{r}
sessionInfo()
```
